### 2018-05-10 manuscript.rmd
updated to bookdown

---
bibliography: manuscript.bib
csl: molecular-phylogenetics-and-evolution.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: mystyles.sty
  word_document: default
---

@@

Siphonophores (Fig. 1 and 2) are among the most abundant gelatinous predators in the open ocean, and have a large impact on ocean ecosystems [@pugh1984diel;@pugh1997vertical; @williams1981vertical; @purcell1981dietary; @pages2001gelatinous;  @Choy20172116]. Siphonophores, which belong to Hydrozoa (Cnidaria), are found at most depths in the ocean, with the deepest recorded species found around 4300m [@lindsay2005planktonic]. The most familiar species is the Portuguese man of war *Physalia physalis*, which floats at the surface and can wash up conspicuously onto beaches [@Totton1960]. Most species are planktonic, living in the water column, where some grow to be more than 30 meters in length [@mackie1987siphonophore]. There is also a small clade of benthic siphonophores, Rhodaliidae, that are tethered to the bottom for part of their lives [@pugh1983benthic]. There are currently 187 valid described siphonophore species [@worms].

Siphonophores remain poorly known, in large part because they are fragile and difficult to collect. They have, however, been of great interest for more than 150 years due to their unique structure and development [@mapstone2014global;@mackie1987siphonophore]. Like many other cnidarians, they are colonial: they grow by incomplete asexual reproduction. Each colony arises from a single embryo that forms the protozooid, the first body. One or two growth zones (Fig. 2) then arise and asexually produce other genetically identical zooids that remain attached [@carre1969etude; @carre1991complete; @carre1967; @carre1993ordre]. These zooids are each homologous to a solitary animal, but are physiologically integrated [@totton1965synopsis; @mackie1987siphonophore; @dunn2006evolution]. Siphonophores differ significantly from other colonial animals in their colony structure and development -- their zooids are highly functionally specialized and arranged in precise, repeating, species-specific patterns [@Beklemishev1969;@cartwright2010character]. Zooids are specialized for a range of functions, including feeding, reproduction, or swimming (Fig. 2) [@dunn2006evolution].

![Photographs of living siphonophores. Colored circles correspond to the clades shown in Figure 3 as follows: Cystonectae (A-B), Calycophorae (C-G), Apolemiidae (H), and Clade A within Euphysonectae (I-K). (A) *Rhizophysa eysenhardtii*, scale bar = 1 cm. (B) *Bathyphysa conifera*, scale bar = 2cm. (C) *Hippopodius hippopus*, scale bar = 5 mm. (D) *Kephyes hiulcus*, scale bar = 2 mm. (E) *Desmophyes haematogaster*, scale bar = 5 mm. (F) *Sphaeronectes christiansonae*, scale bar = 2 mm. (G) *Praya dubia*, scale bar = 4 cm . (H) *Apolemia* sp., scale bar = 1 cm. (I) *Lychnagalma utricularia*, scale bar = 1 cm.  (J) *Nanomia* sp., scale bar = 1 cm. (K) *Physophora hydrostatica*, scale bar = 5 mm. Photo credits: S. Siebert (C,H,I,K), S. Haddock (A,D,E,F), R. Sherlock (B), MBARI (G), C. Munro (J)](figures/Figure1.jpg)

@@

Major questions remained after this early work, though. In particular, there was little support for important deep relationships within Codonophora. Understanding these relationships is key to resolving the evolution of several traits of importance, including sexual systems (monoecy versus dioecy) and the gain and loss of particular zooids, such as palpons (Fig. 2). Here we present a broadly sampled phylogenetic analysis of Siphonophora that considers transcriptomic data from 33 siphonophore species and 10 outgroup species (2 outgroups were subsequently excluded due to poor sampling). Using 1,423 genes, we find strong support for many relationships found in the earlier phylogeny [@Dunn:2005dy], and also provide new resolution for key relationships that were unresolved in that previous study. Using this phylogeny, we reconstruct the evolutionary history of characters central to the unique biology of siphonophores, including zooid type, life history traits, and habitat.

![Schematic of the siphonophore *Nanomia bijuga*, oriented with the anterior of the colony at the top of the page, and the ventral side to the left. Adapted from http://commons.wikimedia.org/wiki/File:Nanomia_bijuga_whole_animal_and_growth_zones.svg, drawn by Freya Goetz. (A) Overview of the whole mature colony. (B) Inset of the pneumatophore and nectosomal growth zone. A series of buds give rise to nectophores. (C) Inset of the siphosomal growth zone. Probuds subdivide to give rise to zooids in repeating units (cormidia). The gastrozooid (specialized feeding polyp) is the posterior-most zooid within each cormidium.](figures/Figure2.pdf)

@@

Specimens were collected in the north-eastern Pacific Ocean, Mediterranean, and the Gulf of California (see table 1). Collection data on all examined specimens, a description of the tissue that was sampled from the colony, collection mode, sample processing details, mRNA extraction methods, sequencing library preparation methods, and sequencing details are summarized in the file Supplementary data 1 (also found in the git repository). Monterey Bay and Gulf of California specimens were collected by remotely operated underwater vehicle (ROV) or during blue-water SCUBA dives. *Chelophyes appendiculata* and *Hippopodius hippopus* (Fig. 1C) specimens were collected in the bay of Villefranche-sur-Mer, France, during a plankton trawl on 13 April 2011. Available physical vouchers have been deposited at the Museum of Comparative Zoology (Harvard University), Cambridge, MA, the Peabody Museum of Natural History (Yale University), New Haven, CT, or had been previously deposited at the Smithsonian National Museum of Natural History, Washington, DC. Accession numbers are given in Supplementary data 1. In cases where physical vouchers were unavailable we provide photographs to document species identity (see git repository: https://github.com/caseywdunn/siphonophore_phylogeny_2017/tree/master/supplementary_info/photographic_vouchers).

@@

Two outgroup species, *Atolla vanhoeffeni* and *Aegina citrea*, were removed from the final supermatrix due to low gene occupancy (gene sampling of 17.0% and 17.3% respectively in a 60% occupancy matrix with 3,379 genes). The final analyses presented here consider 33 siphonophore species and 8 outgroup species. This includes new data for 30 species. In the final analyses, we sampled 1,423 genes to generate a supermatrix with 80% occupancy and a length of 395,699 amino acids (Fig. S1).

@@

We used the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic] to evaluate two hypotheses (Fig. 3C, S2): (i) "Physonectae" is monophyletic [@totton1965synopsis]; (ii) monoecious species are monophyletic [@Dunn:2005dy]. As the sexual system of *Rudjakovia* sp. is unclear, we carried out two tests of the monophyly of monoecy, one with *Rudjakovia* sp. included as a monoecious species, and one with it considered dioecious (Fig. 3C, S2). We used SOWHAT [@doi:10.1093/sysbio/syv055] dev. version 0.39 (commit fd68ef57) to carry out the SOWH tests in parallel, using the default options and an initial sample size of 100 (analysis code can be found in the git repository). For each hypothesis we defined a topology with a single constrained node that was inconsistent with the most likely topology (Fig. 3). We used a threshold for significance of 0.05 and following the initial 100 samples, we evaluated the confidence interval around the p-value to determine if more samples were necessary.

@@

The phylogenetic relationships recovered in this study received strong support across analysis methods (Fig. 3A), with a couple of localized exceptions (Fig. 3B). The ML results were robust to model selection (Fig. S10). Most clades are consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemiidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,423-gene transcriptome analyses. There is strong support for Pyrostephidae as sister to all other non-apolemiid codonophorans. We provisionally refer here to Pyrostephidae as the clade including *Rudjakovia* sp., although sampling of *Pyrostephos vanhoeffeni* is needed in order to determine if *Rudjakovia* sp. falls within Pyrostephidae or is sister to it. Within the clade that is sister to Pyrostephidae, we find two main clades, Calycophorae and a clade we here name Euphysonectae (Fig. 3A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

Euphysonectae consists of two reciprocally monophyletic groups that we here provisionally refer to as Clade A and Clade B (Fig. 3A). The presence of an involucrum, a fold around the base of the cnidoband [@totton1965synopsis], is a potential synapomorphy for Clade A. Species of Clade A also have a descending mantle canal within the nectophores (Fig. S6), a structure that is also present in some calycophorans. Members of Clade A are also monoecious (Fig. 5). There is not a clear synapomorphy for Clade B. Within Clade B there is low support for the placement of *Erenna richardi*, which is placed as sister to Clade A in some analyses (Fig. 3B). More taxon sampling will be required to determine the relationship of species within this clade.

Within Clade A, *Physophora gilmeri* along with *Lychnagalma utricularia* (Fig. 1I) (both not included in the previous phylogeny) are sister to Agalmatidae, a clade restricted to *Agalma*, *Athorybia*, *Melophysa*, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *P. hydrostatica* (the presumed sister species to *P. gilmeri*) was sister to Forskaliidae with low support. The position of *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of Agalmatidae. Placement of *Cordagalma* outside Agalmatidae is consistent with previous analyses of molecular and morphological data [@pugh2006taxonomic; @Dunn:2005dy].

Within Calycophorae, taxon sampling is less comprehensive here than in the previous study. The calycophoran relationships that can be investigated, however, are in broad agreement with the previous analysis. Calycophorans have in the past been split into two groups, prayomorphs and diphyomorphs, based on morphology after @mackie1987siphonophore. As in the previous study, the results presented here indicate that the prayomorphs are paraphyletic with respect to the diphyomorphs. *Craseoa lathetica* and *Desmophyes* sp. are sister to *Hippopodius hippopus* in this study, while in the previous study, the relationship between *C. lathetica* and the clade including *H. hippopus* was unresolved.

![(A) Maximum likelihood (ML) phylogram with bipartition frequencies from the ML bootstraps and the Bayesian posterior distribution of trees. Unlabeled nodes have support >0.99 for both bootstraps and posteriors. The numbers of valid described species estimated to be in each clade based on taxonomy are shown below each clade name on the right. (B) The topologies found in the posterior distribution of trees that conflict with the ML tree. (C) The topologies evaluated by the SOWH tests. For more details on the SOWH topologies refer to Fig. S2.](figures/Figure3.pdf)

Using the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic], we evaluated the following two alternative phylogenetic hypotheses against the most likely tree topology (Fig. 3C): (i) monophyletic Physonectae, (ii) monophyletic monoecious siphonophores. As the sexual mode of  *Rudjakovia* is undescribed, we carried out two tests: with one test including *Rudjakovia* sp. as a monoecious species, and the other with dioecious *Rudjakovia* sp. In all three tests the alternative hypothesis was rejected (p-value <0.01, confidence interval: <0.001 - 0.03, Fig. S2).

The broad taxon sampling and more extensive gene sampling of this phylogeny provide new evidence for the relationships between major siphonophore clades within Codonophora, specifically between Pyrostephidae, Calycophorae, and the newly named Euphysonectae. This opens up new questions about key relationships within both Calycophorae and Euphysonectae -- where future transcriptome sampling efforts should be focused. Within Euphysonectae, two clades (Clade A and Clade B) are hypothesized, although there is weaker support for Clade B (Fig. 3A, 3B). Expanding sampling of species that probably fall in Clade B, including other *Erenna* species, rhodaliids, and relatives of Undescribed sp L, will greatly expand our understanding of these two groups and perhaps provide evidence of Clade B synapomorphies. Similarly, within Calycophorae, increased taxon sampling is needed. This study, and the previous phylogenetic study [@Dunn:2005dy], suggest that the prayomorphs are paraphyletic, but for slightly different reasons given the different sampling of the analyses. In @Dunn:2005dy, a clade of prayomorphs including *Praya dubia* (Fig. 1G), *Nectadamas diomedeae*, and *Nectopyramis natans* (not included in this study) were found to be sister to all other calycophorans, while in this study, the prayomorph *Lilyopsis fluoracantha* (not included in the previous study) is found in a clade including diphyomorph calycophorans that is sister to all other prayomorphs. Expanded transcriptome sequencing, particularly *P. dubia* or a nectopyramid, but also extensive sampling across the major prayomorph and diphyomorph groups, will expand our understanding of relationships within Calycophorae. This will be especially important for understanding trait evolution within Calycophorae,  for example, the release of eudoxids (Fig. 4), or the arrangement of male and female zooids along the stem (see section 3.2 below).

@@

In all siphonophores, each gonophore (sexual medusa that produces gametes) is either male or female. Within each siphonophore species, colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analysis suggested that the common ancestor of siphonophores was dioecious, and was consistent with a single gain of monoecy within Codonophora and no secondary losses [@Dunn:2005dy]. The better-resolved tree in the current analyses indicates that the evolution of monoecy is more complicated than this. The two clades of monoecious siphonophores, Calycophorae and Clade A (Fig. 3A), do not form a monophyletic group. This is because Clade B, which contains dioecious species, is also descended from their most recent common ancestor. SOWH tests strongly reject the placement of the monecious clades Calycophorae and Clade A as a group that excludes Clade B (Figs. 3C and S2). The positions of the only two taxa from Clade B that were included in the previous analysis [@Dunn:2005dy], *Erenna* and *Stephalia*, were unresolved in that study. This difference in conclusions regarding trait evolution, therefore, does not reflect a contradiction between alternative strongly supported results, but the resolution of earlier polytomies in a way that indicates there has been homoplasy in the evolution of monoecy.

The distribution of monoecy is consistent with two potential scenarios (Fig. 4). In the first, there is a single shift from dioecy to monoecy along the branch that gave rise to the most recent common ancestor of Calycophorae and Euphysonectae, followed by a shift back to dioecy along the branch that gave rise to Clade B. In the second, monoecy arose twice: once along the branch that gave rise to Clade A and again along the branch that gave rise to Calycophorae.

Ancestral character state reconstructions favor the hypothesis that monoecy arose twice (Fig. 5A), once in Calycophorae and once in Clade A. This is consistent with differences in the arrangements of male and female gonophores in the two clades. In Clade A, male and female zooids are found within the same cormidium (a single reiterated sequence of zooids along the stem, see Fig. 2). In these species, the male and female zooids are placed at different but well defined locations within the cormidium. Meanwhile in calycophorans, each cormidium bears either male or female gonophores. In this form of monoecy, the male and female cormidia can either occur in an alternating pattern, or there can be several male or female cormidia in a row. In either case, male and female zooids are found at the same corresponding locations within the cormidia. One exception to this is Abylids, where both male or female gonophores may be found within the same eudoxid [@carre1967]. In sum, homoplasy in sexual-system evolution along with variation in the spatial arrangement of gonophores within a colony suggest that siphonophores have evolved different ways to be monoecious. The sexual system and cormidial arrangement of *Rudjakovia* is undescribed, although preliminary observations suggest that this species may be monoecious and that monoecy arose a third time in the Pyrostephidae. A detailed redescription of *Rudjakovia* would help clarify this.

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969; @cartwright2010character]. For example, *Forskalia* and other physonects have at least 5 basic zooid types (nectophore, gastrozooid, palpon, bract, and gonophore), and in some species, there can be nine zooid subtypes (4 types of bract, male & female gonophores)[@pugh2003revision]. Here we reconstruct the evolutionary origins of several zooid types on the present transcriptome-based tree (Fig. 4).

Nectophores (Fig. 2) are non-reproductive propulsive medusae. In Codonophora, the nectophores are localized to a region known as the nectosome (Fig. 2B), which has its own growth zone, and they are used for coordinated colony-level swimming. Planktonic cystonects like *Bathyphysa sibogae* and *Rhizophysa filiformis* (Fig. 1A) instead move through the water column using repeated contraction and relaxation of the stem, and use a gas-filled float and modified flattened gastrozooids with wings (called ptera) to increase surface area and prevent colony sinking [@biggs1976siphonophore].  Nectophores are also present within the gonodendra (reproductive structures) of cystonects, and are thought to propel the gonodendra when they detach from the colony [@totton1965synopsis;@Totton1960]. It is not clear whether the nectophores found within the siphosome of the cystonects are homologous to the nectophores borne on the nectosome of codonophorans. Similarly, the homology of the special nectophore associated with gonophores of the calycophoran *Stephanophyes superba* is also unclear [@chun1891canarischen]. In this study, we only consider the evolution of the nectosome, and not the presence/absence of nectophores. The present analyses, as well as the analyses of @Dunn:2005dy, are consistent with a single origin of the nectosome (Fig. S5).

>>>>

---
bibliography: manuscript.bib
csl: molecular-phylogenetics-and-evolution.csl
site: "bookdown::bookdown_site"
output:
  bookdown::pdf_document2:
    toc: FALSE
---

@@

Siphonophores (Fig. \@ref(siphonophores) and \@ref(colony)) are among the most abundant gelatinous predators in the open ocean, and have a large impact on ocean ecosystems [@pugh1984diel;@pugh1997vertical; @williams1981vertical; @purcell1981dietary; @pages2001gelatinous;  @Choy20172116]. Siphonophores, which belong to Hydrozoa (Cnidaria), are found at most depths in the ocean, with the deepest recorded species found around 4300m [@lindsay2005planktonic]. The most familiar species is the Portuguese man of war *Physalia physalis*, which floats at the surface and can wash up conspicuously onto beaches [@Totton1960]. Most species are planktonic, living in the water column, where some grow to be more than 30 meters in length [@mackie1987siphonophore]. There is also a small clade of benthic siphonophores, Rhodaliidae, that are tethered to the bottom for part of their lives [@pugh1983benthic]. There are currently 187 valid described siphonophore species [@worms].

Siphonophores remain poorly known, in large part because they are fragile and difficult to collect. They have, however, been of great interest for more than 150 years due to their unique structure and development [@mapstone2014global;@mackie1987siphonophore]. Like many other cnidarians, they are colonial: they grow by incomplete asexual reproduction. Each colony arises from a single embryo that forms the protozooid, the first body. One or two growth zones (Fig. \@ref(colony)) then arise and asexually produce other genetically identical zooids that remain attached [@carre1969etude; @carre1991complete; @carre1967; @carre1993ordre]. These zooids are each homologous to a solitary animal, but are physiologically integrated [@totton1965synopsis; @mackie1987siphonophore; @dunn2006evolution]. Siphonophores differ significantly from other colonial animals in their colony structure and development -- their zooids are highly functionally specialized and arranged in precise, repeating, species-specific patterns [@Beklemishev1969;@cartwright2010character]. Zooids are specialized for a range of functions, including feeding, reproduction, or swimming (Fig. \@ref(colony)) [@dunn2006evolution].

![\label{siphonophores}Photographs of living siphonophores. Colored circles correspond to the clades shown in Figure 3 as follows: Cystonectae (A-B), Calycophorae (C-G), Apolemiidae (H), and Clade A within Euphysonectae (I-K). (A) *Rhizophysa eysenhardtii*, scale bar = 1 cm. (B) *Bathyphysa conifera*, scale bar = 2cm. (C) *Hippopodius hippopus*, scale bar = 5 mm. (D) *Kephyes hiulcus*, scale bar = 2 mm. (E) *Desmophyes haematogaster*, scale bar = 5 mm. (F) *Sphaeronectes christiansonae*, scale bar = 2 mm. (G) *Praya dubia*, scale bar = 4 cm . (H) *Apolemia* sp., scale bar = 1 cm. (I) *Lychnagalma utricularia*, scale bar = 1 cm.  (J) *Nanomia* sp., scale bar = 1 cm. (K) *Physophora hydrostatica*, scale bar = 5 mm. Photo credits: S. Siebert (C,H,I,K), S. Haddock (A,D,E,F), R. Sherlock (B), MBARI (G), C. Munro (J)](figures/Figure1.jpg)

@@

Major questions remained after this early work, though. In particular, there was little support for important deep relationships within Codonophora. Understanding these relationships is key to resolving the evolution of several traits of importance, including sexual systems (monoecy versus dioecy) and the gain and loss of particular zooids, such as palpons (Fig. \@ref(colony)). Here we present a broadly sampled phylogenetic analysis of Siphonophora that considers transcriptomic data from 33 siphonophore species and 10 outgroup species (2 outgroups were subsequently excluded due to poor sampling). Using 1,423 genes, we find strong support for many relationships found in the earlier phylogeny [@Dunn:2005dy], and also provide new resolution for key relationships that were unresolved in that previous study. Using this phylogeny, we reconstruct the evolutionary history of characters central to the unique biology of siphonophores, including zooid type, life history traits, and habitat.

![\label{colony}Schematic of the siphonophore *Nanomia bijuga*, oriented with the anterior of the colony at the top of the page, and the ventral side to the left. Adapted from http://commons.wikimedia.org/wiki/File:Nanomia_bijuga_whole_animal_and_growth_zones.svg, drawn by Freya Goetz. (A) Overview of the whole mature colony. (B) Inset of the pneumatophore and nectosomal growth zone. A series of buds give rise to nectophores. (C) Inset of the siphosomal growth zone. Probuds subdivide to give rise to zooids in repeating units (cormidia). The gastrozooid (specialized feeding polyp) is the posterior-most zooid within each cormidium.](figures/Figure2.pdf)

@@

Specimens were collected in the north-eastern Pacific Ocean, Mediterranean, and the Gulf of California (see table 1). Collection data on all examined specimens, a description of the tissue that was sampled from the colony, collection mode, sample processing details, mRNA extraction methods, sequencing library preparation methods, and sequencing details are summarized in the file Supplementary data 1 (also found in the git repository). Monterey Bay and Gulf of California specimens were collected by remotely operated underwater vehicle (ROV) or during blue-water SCUBA dives. *Chelophyes appendiculata* and *Hippopodius hippopus* (Fig. \@ref(siphonophores)C) specimens were collected in the bay of Villefranche-sur-Mer, France, during a plankton trawl on 13 April 2011. Available physical vouchers have been deposited at the Museum of Comparative Zoology (Harvard University), Cambridge, MA, the Peabody Museum of Natural History (Yale University), New Haven, CT, or had been previously deposited at the Smithsonian National Museum of Natural History, Washington, DC. Accession numbers are given in Supplementary data 1. In cases where physical vouchers were unavailable we provide photographs to document species identity (see git repository: https://github.com/caseywdunn/siphonophore_phylogeny_2017/tree/master/supplementary_info/photographic_vouchers).

@@

Two outgroup species, *Atolla vanhoeffeni* and *Aegina citrea*, were removed from the final supermatrix due to low gene occupancy (gene sampling of 17.0% and 17.3% respectively in a 60% occupancy matrix with 3,379 genes). The final analyses presented here consider 33 siphonophore species and 8 outgroup species. This includes new data for 30 species. In the final analyses, we sampled 1,423 genes to generate a supermatrix with 80% occupancy and a length of 395,699 amino acids (Fig. \@ref(gene-occupancy)).

@@

We used the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic] to evaluate two hypotheses (Fig. \@ref(phylogeny)C, S2): (i) "Physonectae" is monophyletic [@totton1965synopsis]; (ii) monoecious species are monophyletic [@Dunn:2005dy]. As the sexual system of *Rudjakovia* sp. is unclear, we carried out two tests of the monophyly of monoecy, one with *Rudjakovia* sp. included as a monoecious species, and one with it considered dioecious (Fig. \@ref(phylogeny)C, S2). We used SOWHAT [@doi:10.1093/sysbio/syv055] dev. version 0.39 (commit fd68ef57) to carry out the SOWH tests in parallel, using the default options and an initial sample size of 100 (analysis code can be found in the git repository). For each hypothesis we defined a topology with a single constrained node that was inconsistent with the most likely topology (Fig. \@ref(phylogeny)). We used a threshold for significance of 0.05 and following the initial 100 samples, we evaluated the confidence interval around the p-value to determine if more samples were necessary.

@@

The phylogenetic relationships recovered in this study received strong support across analysis methods (Fig. \@ref(phylogeny)A), with a couple of localized exceptions (Fig. \@ref(phylogeny)B). The ML results were robust to model selection (Fig. \@ref(fig:supplementary-trees)). Most clades are consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemiidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,423-gene transcriptome analyses. There is strong support for Pyrostephidae as sister to all other non-apolemiid codonophorans. We provisionally refer here to Pyrostephidae as the clade including *Rudjakovia* sp., although sampling of *Pyrostephos vanhoeffeni* is needed in order to determine if *Rudjakovia* sp. falls within Pyrostephidae or is sister to it. Within the clade that is sister to Pyrostephidae, we find two main clades, Calycophorae and a clade we here name Euphysonectae (Fig. \@ref(phylogeny)A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

In ML analyses Euphysonectae consists of two reciprocally monophyletic groups that we here provisionally refer to as Clade A and Group B (Fig. \@ref(phylogeny)A). In BI analyses, Group B is graded (Fig. \@ref(phylogeny)B). The presence of an involucrum, a fold around the base of the cnidoband [@totton1965synopsis], is a potential synapomorphy for Clade A. Species of Clade A also have a descending mantle canal within the nectophores (Fig. \@ref(fig:desc-mantle), \@ref(fig:mantle-BI)), a structure that is also present in some calycophorans. Members of Clade A are also monoecious (Fig. \@ref(reconstruct)). There is not a clear synapomorphy for Group B. Within Group B there is high support for the placement of *Erenna richardi* in ML analyses, but it is placed as sister to Clade A in BI analyses (Fig. \@ref(phylogeny)B). More taxon sampling will be required to determine the relationship of species within this clade.

Within Clade A, *Physophora gilmeri* along with *Lychnagalma utricularia* (Fig. \@ref(siphonophores)I) (both not included in the previous phylogeny) are sister to Agalmatidae, a clade restricted to *Agalma*, *Athorybia*, *Melophysa*, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *P. hydrostatica* (the presumed sister species to *P. gilmeri*) was sister to Forskaliidae with low support. The position of *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of Agalmatidae. Placement of *Cordagalma* outside Agalmatidae is consistent with previous analyses of molecular and morphological data [@pugh2006taxonomic; @Dunn:2005dy].

Within Calycophorae, taxon sampling is less comprehensive here than in the previous study. The calycophoran relationships that can be investigated, however, are in broad agreement with the previous analysis. Calycophorans have in the past been split into two groups, prayomorphs and diphyomorphs, based on morphology after @mackie1987siphonophore. As in the previous study, the results presented here indicate that the prayomorphs are paraphyletic with respect to the diphyomorphs. In the previous study, the relationship between *C. lathetica* and the clade including *H. hippopus* was unresolved. In this study, *Craseoa lathetica* and *Desmophyes* sp. are sister to *Hippopodius hippopus* in ML analyses with high support, while in BI analyses, *H. hippopus* is sister to *Lilyopsis fluoracantha* and the diphyomorphs (Fig. \@ref(phylogeny)B).

![\label{phylogeny}(A) Maximum likelihood (ML) phylogram with bipartition frequencies from the ML bootstraps and the Bayesian posterior distribution of trees. Unlabeled nodes have support >0.99 for both bootstraps and posteriors. The numbers of valid described species estimated to be in each clade based on taxonomy are shown below each clade name on the right. (B) The topologies found in the posterior distribution of trees that conflict with the ML tree. (C) The topologies evaluated by the SOWH tests. For more details on the SOWH topologies refer to Fig. \@ref(sowh).](figures/Figure3.pdf)

Using the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic], we evaluated the following two alternative phylogenetic hypotheses against the most likely tree topology (Fig. \@ref(phylogeny)C): (i) monophyletic Physonectae, (ii) monophyletic monoecious siphonophores. As the sexual mode of  *Rudjakovia* is undescribed, we carried out two tests: with one test including *Rudjakovia* sp. as a monoecious species, and the other with dioecious *Rudjakovia* sp. In all three tests the alternative hypothesis was rejected (p-value <0.01, confidence interval: <0.001 - 0.03, Fig. \@ref(sowh)).

The broad taxon sampling and more extensive gene sampling of this phylogeny provide new evidence for the relationships between major siphonophore clades within Codonophora, specifically between Pyrostephidae, Calycophorae, and the newly named Euphysonectae. This opens up new questions about key relationships within both Calycophorae and Euphysonectae -- where future transcriptome sampling efforts should be focused. Within Euphysonectae, two clades (Clade A and Group B) are hypothesized, although there is weaker support for Group B (Fig. \@ref(phylogeny)A, 3B). Expanding sampling of species that probably fall in Group B, including other *Erenna* species, rhodaliids, and relatives of Undescribed sp L, will greatly expand our understanding of these two groups and perhaps provide evidence of Group B synapomorphies. Similarly, within Calycophorae, increased taxon sampling is needed. This study, and the previous phylogenetic study [@Dunn:2005dy], suggest that the prayomorphs are paraphyletic, but for slightly different reasons given the different sampling of the analyses. In @Dunn:2005dy, a clade of prayomorphs including *Praya dubia* (Fig. \@ref(siphonophores)G), *Nectadamas diomedeae*, and *Nectopyramis natans* (not included in this study) were found to be sister to all other calycophorans, while in this study, the prayomorph *Lilyopsis fluoracantha* (not included in the previous study) is found in a clade including diphyomorph calycophorans that is sister to all other prayomorphs. Expanded transcriptome sequencing, particularly *P. dubia* or a nectopyramid, but also extensive sampling across the major prayomorph and diphyomorph groups, will expand our understanding of relationships within Calycophorae. This will be especially important for understanding trait evolution within Calycophorae,  for example, the release of eudoxids (Fig. \@ref(traitdist)), or the arrangement of male and female zooids along the stem (see section 3.2 below).

@@

In all siphonophores, each gonophore (sexual medusa that produces gametes) is either male or female. Within each siphonophore species, colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analysis suggested that the common ancestor of siphonophores was dioecious, and was consistent with a single gain of monoecy within Codonophora and no secondary losses [@Dunn:2005dy]. The better-resolved tree in the current analyses indicates that the evolution of monoecy is more complicated than this. The two clades of monoecious siphonophores, Calycophorae and Clade A (Fig. \@ref(phylogeny)A), do not form a monophyletic group. This is because Group B, which contains dioecious species, is also descended from their most recent common ancestor. SOWH tests strongly reject the placement of the monecious clades Calycophorae and Clade A as a group that excludes Group B (Figs. 3C and S2). The positions of the only two taxa from Group B that were included in the previous analysis [@Dunn:2005dy], *Erenna* and *Stephalia*, were unresolved in that study. This difference in conclusions regarding trait evolution, therefore, does not reflect a contradiction between alternative strongly supported results, but the resolution of earlier polytomies in a way that indicates there has been homoplasy in the evolution of monoecy.

The distribution of monoecy is consistent with two potential scenarios (Fig. \@ref(traitdist)). In the first, there is a single shift from dioecy to monoecy along the branch that gave rise to the most recent common ancestor of Calycophorae and Euphysonectae, followed by a shift back to dioecy along the branch that gave rise to Group B. In the second, monoecy arose twice: once along the branch that gave rise to Clade A and again along the branch that gave rise to Calycophorae.

Ancestral character state reconstructions favor the hypothesis that monoecy arose twice (Fig. \@ref(reconstruct)A, \@ref(fig:BI-mono)), once in Calycophorae and once in Clade A. This is consistent with differences in the arrangements of male and female gonophores in the two clades. In Clade A, male and female zooids are found within the same cormidium (a single reiterated sequence of zooids along the stem, see Fig. \@ref(colony)). In these species, the male and female zooids are placed at different but well defined locations within the cormidium. Meanwhile in calycophorans, each cormidium bears either male or female gonophores. In this form of monoecy, the male and female cormidia can either occur in an alternating pattern, or there can be several male or female cormidia in a row. In either case, male and female zooids are found at the same corresponding locations within the cormidia. One exception to this is Abylids, where both male or female gonophores may be found within the same eudoxid [@carre1967]. In sum, homoplasy in sexual-system evolution along with variation in the spatial arrangement of gonophores within a colony suggest that siphonophores have evolved different ways to be monoecious. The sexual system and cormidial arrangement of *Rudjakovia* is undescribed, although preliminary observations suggest that this species may be monoecious and that monoecy arose a third time in the Pyrostephidae. A detailed redescription of *Rudjakovia* would help clarify this.

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969; @cartwright2010character]. For example, *Forskalia* and other physonects have at least 5 basic zooid types (nectophore, gastrozooid, palpon, bract, and gonophore), and in some species, there can be nine zooid subtypes (4 types of bract, male & female gonophores)[@pugh2003revision]. Here we reconstruct the evolutionary origins of several zooid types on the present transcriptome-based tree (Fig. \@ref(traitdist)).

Nectophores (Fig. \@ref(colony)) are non-reproductive propulsive medusae. In Codonophora, the nectophores are localized to a region known as the nectosome (Fig. \@ref(colony)B), which has its own growth zone, and they are used for coordinated colony-level swimming. Planktonic cystonects like *Bathyphysa sibogae* and *Rhizophysa filiformis* (Fig. \@ref(siphonophores)A) instead move through the water column using repeated contraction and relaxation of the stem, and use a gas-filled float and modified flattened gastrozooids with wings (called ptera) to increase surface area and prevent colony sinking [@biggs1976siphonophore].  Nectophores are also present within the gonodendra (reproductive structures) of cystonects, and are thought to propel the gonodendra when they detach from the colony [@totton1965synopsis;@Totton1960]. It is not clear whether the nectophores found within the siphosome of the cystonects are homologous to the nectophores borne on the nectosome of codonophorans. Similarly, the homology of the special nectophore associated with gonophores of the calycophoran *Stephanophyes superba* is also unclear [@chun1891canarischen]. In this study, we only consider the evolution of the nectosome, and not the presence/absence of nectophores. The present analyses, as well as the analyses of @Dunn:2005dy, are consistent with a single origin of the nectosome (Fig. \@ref(fig:nectosome-origin), \@ref(fig:nectosome-BI)).
